Sequence Specific Modeling of Cell-Free Protein Synthesis

### 2016-12-22 Paper_v1.tex
edits and supplement

\usepackage{natbib}
\usepackage{setspace}
\usepackage{xkeyval}
\usepackage{url}  % Formatting web addresses
\usepackage{ifthen}  % Conditional
\usepackage{multicol}   %Columns
\usepackage[utf8]{inputenc} %unicode support
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{epsfig}
\usepackage{epstopdf}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
\usepackage[version=3]{mhchem} % Formula subscripts using \ce{}
\usepackage[T1]{fontenc}       % Use modern font encodings
%\usepackage[square,sort,comma,numbers,sort&compress]{natbib}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% If issues arise when submitting your manuscript, you may want to
%% un-comment the next line.  This provides information on the
%% version of every file you have used.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%\listfiles

\bibliographystyle{achemso}
\newcommand*\mycommand[1]{\texttt{\emph{#1}}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Meta-data block
%% ---------------
%% Each author should be given as a separate \author command.
%%
%% Corresponding authors should have an e-mail given after the author
%% name as an \email command. Phone and fax numbers can be given
%% using \phone and \fax, respectively; this information is optional.
%%
%% The affiliation of authors is given after the authors; each
%% \affiliation command applies to all preceding authors not already
%% assigned an affiliation.
%%
%% The affiliation takes an option argument for the short name.  This
%% will typically be something like "University of Somewhere".
%%
%% The \altaffiliation macro should be used for new address, etc.
%% On the other hand, \alsoaffiliation is used on a per author basis
%% when authors are associated with multiple institutions.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author{Michael Vilkhovoy}
\author{Nicholas Horvath}
\author{Jeffrey D. Varner}
\email{jdv27@cornell.edu}
\phone{+1~(607)~255-4258}
\fax{+1~(607)~255-9166}
\affiliation[Cornell University]
{Robert Frederick Smith School of Chemical and Biomolecular Engineering, Cornell University, Ithaca, NY 14853}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Sequence Specific Modeling of Cell-Free Protein Synthesis}
%\abbreviations{MV,NH,JDV}
\keywords{Synthetic biology, Constraints based modeling, Biochemical modeling}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The manuscript does not need to include \maketitle, which is
%% executed automatically.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The "tocentry" environment can be used to create an entry for the
%% graphical table of contents. It is given here as some journals
%% require that it is printed as part of the abstract page. It will
%% be automatically moved as appropriate.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Start the main part of the manuscript here.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
Cell free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
Central amongst these advantages is direct access to metabolites and the microbial biosynthetic machinery without the interference of a cell wall.
This allows us to control as well as interrogate the chemical environment while the biosynthetic machinery is operating, potentially at a fine time resolution.
Second, cell-free systems also allow us to study biological processes without the complications associated with cell growth.
Cell-free protein synthesis (CFPS) systems are arguably the most prominent examples of cell-free systems used today \cite{Jewett:2008aa}.
However, CFPS is not new; CFPS in crude \textit{E.~coli} extracts has been used since the 1960s to explore fundamentally important biological mechanisms \cite{MATTHAEI:1961aa,NIRENBERG:1961aa}.
Today, cell-free systems are used in a variety of applications ranging from therapeutic protein production \cite{Lu:2014aa} to synthetic biology \cite{Hodgman:2012aa}.
Interestingly, many of the challenges confronting in-vivo genome-scale kinetic modeling can potentially be overcome in a cell-free system.
For example, there is no complex transcriptional regulation to consider, transient metabolic measurements are easier to obtain, and we no longer have to consider cell growth.
Thus, cell-free operation holds several significant advantages for model development, identification and validation.
Theoretically, genome-scale cell-free kinetic models may be possible for industrially important organisms, such as \textit{E. coli} or \textit{B. subtilis}, if a simple, tractable framework for integrating allosteric regulation with enzyme kinetics can be formulated.

@@

In this study, we used sequence specific constraints based modeling to evaluate the performance of synthetic circuits in an \emph{E.~coli} cell-free protein synthesis (CFPS) system.
A core \emph{E.~coli} cell free metabolic model, consisting of 132 metabolites and 281 reactions, was developed from literature \cite{Feist:2007aa}.
This model, which described glycolysis, pentose phosphate pathway, amino acid biosynthesis and degradation and energy metabolism, was then augmented with
sequence specific descriptions of genetic circuits which included mechanistic models of promoter function, transcription and translation.
Thus, sequence specific constraints based modeling explicitly couples the transcription and translation of circuit components with the availability of metabolic resources.
We tested this approach by first simulating $\sigma_{70}$-induced deGFP expression at different plasmid concentrations and then expanded this for the production of CAT and deGFP under two different promoters.
First principles predictions of circuit performance were consistent with measurements.
We then investigated the productivity and carbon yield for a range of 8 different proteins for a variety of cases.
From this, higher carbon number proteins typically had lower productivity rates and carbon yields than that of the lower carbon number proteins.
Further, global sensitivity analysis identified the key metabolic processes that controlled circuit performance, showing oxidative phosphorylation is instrumental for maintaining a high carbon yield and the translation rate for productivity.
Taken together, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell free synthetic circuits.

@@

\subsection{Model Derivation}
The goal of this work was first to construct a modeling framework to describe cell-free protein synthesis systems and to examine its performance in productivity and yield for a protein of interest.
One mathematical framework that has found wide use in modeling metabolism are constraint based models such as flux balance analysis (FBA).
FBA can predict how cells utilize nutrients to produce products by using the biochemical stoichiometry and thermodynamical feasibility under pseudo steady-state conditions.
Traditionally, FBA is used to model \textit{in vivo} processes, however cell-free systems do not have growth associated reactions or transport through the cell membrane.
Thus, to model cell-free metabolism we constructed a cell-free stoichiometric network by removing growth associated reactions from the MG1655 reconstruction \cite{Feist:2007aa}, and incorporating amino acid synthesis and transcription/translation associated reactions \cite{Allen:2003aa} for a protein of interest to be expressed.
The network consisted of 281 reactions and 132 species.
We developed a cell-free sequence specific flux balance analysis (ssFBA) with a detailed promoter model \cite{Moon:2012aa} to examine the performance of CFPS.
We first validated the ssFBA approach by comparing simulated and measured concentrations of two proteins from two different cell-free \textit{E. coli} extract systems.
The first protein, chloramphenicol acetyltransferase (CAT), was produced under a T7 promoter in a glucose/NMP cell-free system \cite{2005_calhoun_BiotechnologyProgress} for a duration of 1 hour under glucose consumption (Fig \ref{fig:CAT_GFP_prod}A).
The second protein, dual emission green fluorescent protein (deGFP), was produced under a P70 promoter in TXTL 2.0 \textit{E. coli} extract for a duration of 8 hours under maltose consumption(Fig \ref{fig:CAT_GFP_prod}B).
The ssFBA simulations predicted the production of both proteins for the duration of both CFPS batch reactions.
Uncertainty in experimental factors such as RNA polymerase, ribosome concentrations, elongation rates or the upper bounds for oxygen and glucose consumption rates did not alter the qualitative performance of the model.
Thus, the metabolic network and molecular description of trancription and translation were consistent with experimental measurements.
\begin{figure}[t!]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/CAT_GFP_prod.pdf}
\caption{Sequence specific flux balance analysis of protein production vesus time. A. CAT production under a T7 promoter in CFPS \textit{E. coli} extract for 1 hr under glucose consumption. B. deGFP production under a P70 promoter in TXTL 2.0 \textit{E. coli} extract for 8 hr under glucose consumption. 95\% CI (blue region) over the ensemble of 100 sets.}
\label{fig:CAT_GFP_prod}
\end{figure}

Next, ssFBA predicted deGFP production as a function of plasmid concentration (Fig \ref{fig:GFP_plasmid}).
Concentration of deGFP at each plasmid concentration was calculated by multiplying the flux of deGFP synthesis by the active time of production, approximately 8 hours in TXTL 2.0 \cite{2016_Garamella_etal_TXTL}.
The mean of the ensemble shows a good prediction against the measured deGFP levels, even though it under predicted deGFP concentration at the saturating point of 5 nM of plasmid concentration.
However, the ensemble and the mean of the ensemble captured the overall saturating dynamics of deGFP production as a function of plasmid concentration.
\begin{figure}[t!]
\centering
\includegraphics[width=0.6\textwidth]{./Figures/GFP_plasmid.pdf}
\caption{Predicted deGFP concentration (black line) at different plasmid concentrations versus measurements of deGFP (dots) synthesized in TXTL 2.0. 95\% CI (blue region) over the ensemble of 100 sets.}
\label{fig:GFP_plasmid}
\end{figure}

These results validated our mathematical framework to model CFPS systems and predict the production of two proteins with no adjustable parameters.
It also showed that the sequence specific reactions were sufficient to predict the production of two different proteins under different promoters and cell-free systems.
Since the model accurately predicted the behavior of protein production, we wanted to use our mathematical framework to help us understand the performance limits of CFPS and how these shortcomings could be addressed.

\subsection{CFPS Productivity}
Our next goal was to examine the performance of CFPS productivity for eight different proteins under three different cases. Each of the proteins were produced under a P70 promoter, except for CAT which was produced under a T7 promoter. In all our cases, CFPS is supplied with glucose. The first case, CFPS is supplied with amino acids and the system can synthesis amino acids (control). In the second case, CFPS is supplied with amino acids, however the system cannot synthesis amino acids (AA uptake w/o synthesis). We turned off these synthesis reactions because during the cell-free extract preparation the cells are often supplied with amino acids, thus the enzymes responsible for amino acid synthesis would not be present. In the third case, CFPS is not supplied with amino acids, but the system can synthesis them (AA synthesis w/o uptakte).
We used ssFBA to estimate the productivity of eight proteins for each case (Fig \ref{fig:Prod_POI}A).
The second case (without amino acid synthesis) showed the highest productivity for each of the proteins, however the control case had very similar performance.
This shows the system has sufficient substrates to power the system and synthesis each protein.
The third case had the lowest productivity for each protein, thus the addition of amino acids to CFPS extract is important for maintaining a relatively high productivity.
The qualitative trend of productivity for the three different cases was the same, however some proteins had higher productivity than others.
For instance, in the second case FGF21 had a productivity of 17 ($\mu$M/h) whereas FII had a productivity of 3 ($\mu$M/h).
To examine this further, the mean productivity was plotted against the carbon number of each protein (Fig \ref{fig:Prod_POI}B).
The proteins with the highest productivity had the lowest carbon number whereas proteins with low productivity had higher carbon numbers.
This inverse qualitative trend was due to the fact that larger proteins require more amino acids and substrates to assemble them resulting in lower productivity.
\begin{figure}[t!]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/Prod_POI.pdf}
\caption{CFPS productivity performace for eight proteins for the control (black), AA uptake w/o synthesis (off white), and AA synthesis w/o uptake (grey). A. Productivity normalized to the specific glucose uptake rate (Error bars represent the 95\% CI of the ensemble). B. Mean productivity versus the carbon number for each corresponding protein.}
\label{fig:Prod_POI}
\end{figure}

\subsection{CFPS Carbon Yield}
Following the same outline as in examining the productivity, we calculated the carbon yield for each protein (Fig \ref{fig:Yield_POI}A).
The same trends followed, where the case without amino acid synthesis showed the highest carbon yield and the control case had comparable performance.
The third case (with no amino acid uptake) had the lowest yields; this is most likely because glucose is now utilized to synthesize the necessary amino acids for each protein as well as power the system.
We also investigated the carbon contribution of glucose and each amino acid toward the production of deGFP for all three cases (Table \ref{tbl:yield_breakdown}).
For the first case, the system is relying on a mixture of glucose and some amino acids for the productin of deGFP with a carbon yield of 20.5\%.
Once amino acid synthesis is removed from the network (2nd case), each amino acid is utilized and the carbon yield is increased to 22.2\%.
Only the necessary amount of amino acids are used for the production of CAT, thus it may be hypothesized that all the glucose is used to power CFPS and does not contribute to the carbon yield.
If that is the case, the carbon yield without glucose contribution is 100\%, whereas for the first case its 92.5\%.
Finally for the third case (without amino acid supplemented), the carbon yield reduces to 14.8\% and the system uses about double the amount of glucose as the first two cases.
For this case glucose is being used to synthesize amino acids and power the system and this trend was seen across all proteins except for CAT.

\begin{table}[t!]
\centering
    \caption{Carbon contribution from glucose and each amino acid toward deGFP production for three cases: control, amino acid uptake without synthesis, and amino acid synthesis without uptake.}
    \renewcommand{\arraystretch}{0.95}
    \begin{tabular}{lccc} \toprule
        \parbox{3.2 cm}{Carbon \phantom{abcdefghi} Produced (mM)} & Control & \parbox{2.5 cm}{AA uptake \phantom{ai} w/o synthesis} & \parbox{2.5 cm}{AA synthesis w/o uptake} \\ \midrule
        deGFP & 9.8 & 9.6 & 9.9  \\ \toprule
        \parbox{3.2 cm}{Carbon \phantom{abcdefghi} Consumed (mM)} & \\ \midrule
        GLC & 37.3 & 33.7 & 66.9 \\
        ALA & 0.0 & 0.2 & -  \\
        ARG & 0.3 & 0.3 & - \\
        ASN & 0.5 & 0.4 & -  \\
        ASP & 0.4 & 0.6 & -  \\
        CYS & 0.2 & 0.1 & -  \\
        GLU & 2.1 & 0.6 & -  \\
        GLN & 0.4 & 0.3 & - \\
        GLY & 0.3 & 0.3 & -  \\
        HIS & 0.5 & 0.5 & - \\
        ILE & 0.0 & 0.6 & -  \\
        LEU & 1.0 & 1.0 & - \\
        LYS & 1.0 & 0.9 & - \\
		    MET & 0.0 & 0.2 & -  \\
        PHE & 1.0 & 0.9 & - \\
        PRO & 0.4 & 0.4 & -  \\
        SER & 0.0 & 0.2 & - \\
        THR & 1.0 & 0.5 & -  \\
        TRP & 0.1 & 0.1 & -  \\
        TYR & 0.8 & 0.8 & - \\
        VAL & 0.6 & 0.7 & - \\ \hline
        Sum & 47.9 & 43.3 & 66.9  \\ \toprule
        Yield & 20.5\% & 22.2\% & 14.8\% \\
	Yield w/o GLC & 92.5\% & 100\% & - \\ \bottomrule
    \end{tabular}
\label{tbl:yield_breakdown}
\end{table}

Interestingly, CAT carbon yield showed similar performance for all three cases.
Each protein has different energy requirements for transcription and translation depending on its sequence.
Thus, for the case of CAT, energy requirements for transcription or translation are low enough that it's carbon yield is not hindered in the third case.
We next investigated the effect of the carbon number of each protein to the carbon yield (Fig \ref{fig:Yield_POI}B).
The same inverse qualitative trend is observed, however it is less significant.
Sequence specific flux balance analysis assumes a psuedo steady state, thus intermediate metabolites cannot be accumulated within the cell-free extract.
In addition, ssFBA is solved by setting the production of the protein as the objective function.
Therefore, carbon flux will travel throughout the network to optimize the flux through the protein synthesis reaction.
\begin{figure}[t!]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/Yield_POI.pdf}
\caption{CFPS carbon yield performace for eight proteins for the control (black), AA uptake w/o synthesis (off white), and AA synthesis w/o uptake (grey). A. Carbon yield (Error bars represent the 95\% CI of the ensemble). B. Mean carbon yield versus the carbon number for each corresponding protein.}
\label{fig:Yield_POI}
\end{figure}

>>>>

\usepackage{natbib}
\usepackage{setspace}
\usepackage{xkeyval}
\usepackage{array}
\usepackage{listings}
\usepackage{lmodern}
\usepackage{mathpazo}
\usepackage{geometry}
\usepackage{microtype}
\usepackage{url}  % Formatting web addresses
\usepackage{ifthen}  % Conditional
\usepackage{multicol}   %Columns
\usepackage[utf8]{inputenc} %unicode support
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{epsfig}
\usepackage{epstopdf}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{natmove}
\usepackage{float}
\usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
\usepackage[version=3]{mhchem} % Formula subscripts using \ce{}
\usepackage[T1]{fontenc}       % Use modern font encodings
%\usepackage[square,sort,comma,numbers,sort&compress]{natbib}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% If issues arise when submitting your manuscript, you may want to
%% un-comment the next line.  This provides information on the
%% version of every file you have used.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%\listfiles

%\bibliographystyle{achemso}
\newcommand*\mycommand[1]{\texttt{\emph{#1}}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author{Michael Vilkhovoy}
\author{Nicholas Horvath}
\author{Jeffrey D. Varner}
\email{jdv27@cornell.edu}
\phone{+1~(607)~255-4258}
\fax{+1~(607)~255-9166}
\affiliation[Cornell University]
{Robert Frederick Smith School of Chemical and Biomolecular Engineering, Cornell University, Ithaca, NY 14853}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Sequence Specific Modeling of Cell-Free Protein Synthesis}
%\abbreviations{MV,NH,JDV}
\keywords{Synthetic biology, Constraints based modeling, Biochemical modeling}

\begin{document}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Start the main part of the manuscript here.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}
Cell free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
Central amongst these advantages is direct access to metabolites and the microbial biosynthetic machinery without the interference of a cell wall.
This allows us to control as well as interrogate the chemical environment while the biosynthetic machinery is operating, potentially at a fine time resolution.
Second, cell-free systems also allow us to study biological processes without the complications associated with cell growth.
Cell-free protein synthesis (CFPS) systems are arguably the most prominent examples of cell-free systems used today \cite{Jewett:2008aa}.
However, CFPS is not new; CFPS in crude \textit{E.~coli} extracts has been used since the 1960s to explore fundamentally important biological mechanisms \cite{MATTHAEI:1961aa,NIRENBERG:1961aa}.
Today, cell-free systems are used in a variety of applications ranging from therapeutic protein production \cite{Lu:2014aa} to synthetic biology \cite{Hodgman:2012aa}.
Interestingly, many of the challenges confronting in-vivo genome-scale kinetic modeling can potentially be overcome in a cell-free system.
For example, there is no complex transcriptional regulation to consider, transient metabolic measurements are easier to obtain, and we no longer have to consider cell growth.
Thus, cell-free operation holds several significant advantages for model development, identification and validation.
Theoretically, genome-scale cell-free kinetic models may be possible for industrially important organisms, such as \textit{E. coli} or \textit{B. subtilis}, if a simple, tractable framework for integrating allosteric regulation with enzyme kinetics can be formulated.

@@

In this study, we used sequence specific constraints based modeling to evaluate the performance of synthetic circuits in an \emph{E.~coli} cell-free protein synthesis (CFPS) system.
A core \emph{E.~coli} cell free metabolic model, consisting of 132 metabolites and 281 reactions, was developed from literature \cite{Feist:2007aa}.
This model, which described glycolysis, pentose phosphate pathway, amino acid biosynthesis and degradation and energy metabolism, was then augmented with
sequence specific descriptions of genetic circuits which included mechanistic models of promoter function, transcription and translation.
Thus, sequence specific constraints based modeling explicitly coupled the transcription and translation of circuit components with the availability of metabolic resources.
We tested this approach by first simulating the production of CAT and deGFP under two different promoters. and then expanded this for $\sigma_{70}$-induced deGFP expression at different plasmid concentrations.
First principles predictions of circuit performance were consistent with measurements.
We then investigated the productivity and carbon yield for a range of 8 different proteins for a variety of cases.
From this, higher carbon number proteins typically had lower productivity rates and carbon yields than that of the lower carbon number proteins.
Further, global sensitivity analysis identified the key metabolic processes that controlled circuit performance, showing oxidative phosphorylation is instrumental for maintaining a high carbon yield and the translation rate for productivity.
Taken together, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell free synthetic circuits.

@@

\subsection{Model Derivation}
The goal of this work was first to construct a modeling framework to describe cell-free protein synthesis systems and to examine its performance in productivity and yield for a protein of interest.
One mathematical framework that has found wide use in modeling metabolism are constraint based models such as flux balance analysis (FBA).
FBA can predict how cells utilize nutrients to produce products by using the biochemical stoichiometry and thermodynamical feasibility under pseudo steady-state conditions.
Traditionally, FBA is used to model \textit{in vivo} processes, however cell-free systems do not have growth associated reactions or transport through the cell membrane.
Thus, to model cell-free metabolism we constructed a cell-free stoichiometric network by removing growth associated reactions from the MG1655 reconstruction \cite{Feist:2007aa}, and incorporating amino acid synthesis and transcription/translation associated reactions \cite{Allen:2003aa} for a protein of interest to be expressed.
The network consisted of 281 reactions and 132 species.
We developed a cell-free sequence specific flux balance analysis (ssFBA) with a detailed promoter model \cite{Moon:2012aa} to examine the performance of CFPS.
We first validated the ssFBA approach by comparing simulated and measured concentrations of two proteins from two different cell-free \textit{E. coli} extract systems.
The first protein, chloramphenicol acetyltransferase (CAT), was produced under a T7 promoter in a glucose/NMP cell-free system \cite{2005_calhoun_BiotechnologyProgress} for a duration of 1 hour under glucose consumption (Fig \ref{fig:CAT_GFP_prod}A).
The second protein, dual emission green fluorescent protein (deGFP), was produced under a P70 promoter in TXTL 2.0 \textit{E. coli} extract for a duration of 8 hours under maltose consumption (Fig \ref{fig:CAT_GFP_prod}B).
The ssFBA simulations predicted the production of both proteins for the duration of both CFPS batch reactions.
Uncertainty in experimental factors such as RNA polymerase, ribosome concentrations, elongation rates or the upper bounds for oxygen and glucose consumption rates did not alter the qualitative performance of the model.
Thus, the metabolic network and molecular description of trancription and translation were consistent with experimental measurements.
\begin{figure}[t!]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/CAT_GFP_prod.pdf}
\caption{Sequence specific flux balance analysis of protein production vesus time. A. CAT production under a T7 promoter in CFPS \textit{E. coli} extract for 1 hr under glucose consumption. B. deGFP production under a P70 promoter in TXTL 2.0 \textit{E. coli} extract for 8 hr under glucose consumption. 95\% CI (blue region) over the ensemble of 100 sets.}
\label{fig:CAT_GFP_prod}
\end{figure}

Next, ssFBA predicted deGFP production as a function of plasmid concentration (Fig \ref{fig:GFP_plasmid}).
Concentration of deGFP at each plasmid concentration was calculated by multiplying the flux of deGFP synthesis by the active time of production, approximately 8 hours in TXTL 2.0 \cite{Garamella:2016aa}.
The mean of the ensemble shows a good prediction against the measured deGFP levels, even though it under predicted deGFP concentration at the saturating point of 5 nM of plasmid concentration.
However, the ensemble and the mean of the ensemble captured the overall saturating dynamics of deGFP production as a function of plasmid concentration.
\begin{figure}[t!]
\centering
\includegraphics[width=0.6\textwidth]{./Figures/GFP_plasmid.pdf}
\caption{Predicted deGFP concentration (black line) at different plasmid concentrations versus measurements of deGFP (dots) synthesized in TXTL 2.0. 95\% CI (blue region) over the ensemble of 100 sets.}
\label{fig:GFP_plasmid}
\end{figure}

These results validated our mathematical framework to model CFPS systems and predict the production of two proteins with no adjustable parameters.
It also showed that the sequence specific reactions were sufficient to predict the production of two different proteins under two different promoters and cell-free systems.
Since the model accurately predicted the behavior of protein production, we wanted to use our mathematical framework to help us understand the performance limits of CFPS and how these shortcomings could be addressed.

\subsection{CFPS Productivity}
Our next goal was to examine the performance of CFPS productivity for eight different proteins under three different cases. Each of the proteins were produced under a P70 promoter, except for CAT which was produced under a T7 promoter. In all our cases, CFPS is supplied with glucose. The first case, CFPS is supplied with amino acids and the system can synthesize amino acids (control). In the second case, CFPS is supplied with amino acids, however the system cannot synthesize amino acids (AA uptake w/o synthesis). We turned off these synthesis reactions since during the cell-free extract preparation the cells are often supplied with amino acids, thus the enzymes responsible for amino acid synthesis would not be present. In the third case, CFPS is not supplied with amino acids, but the system can synthesize them (AA synthesis w/o uptakte).
Eight different proteins,ranging in size, were selected to evaluate CFPS performance: bone morphogenetic protein 10 (BMP10), chloramphenicol acetyltransferase (CAT), caspase 9 (CASP9), dual emission green fluorescent protein (deGFP), prothrombin (FII), coagulation factor X (FX), fibroblast growth factor 21 (FGF21), and single chain variable fragment R4 (scFvR4).
We used ssFBA to estimate the productivity for each of these proteins for each case (Fig \ref{fig:Prod_POI}A).
The first two cases had very similar performance, with the second case (without amino acid synthesis) having a slightly higher productivity for most proteins. 
%The second case (without amino acid synthesis) showed the highest productivity for each of the proteins, however the control case had very similar performance.
This shows the system had sufficient substrates to power the system and synthesize each protein.
The third case had the lowest productivity for each protein, thus the addition of amino acids to CFPS extract is important for maintaining a relatively high productivity.
The drop in productivity in the third case is due to the fact that glucose is the only substrate supplied and must be used to provide the necessary energy requirements for transcription and translation as well as synthesize each amino acids required for the production of the protein of interest.  
The qualitative trend of productivity for the three different cases was the same within each case, however some proteins had higher productivity than others.
For instance, in the second case FGF21 (fibroblast growth factor 21) had a productivity of 17 ($\mu$M/h) whereas FII (prothrombin) had a productivity of 3 ($\mu$M/h).
To examine this further, the mean productivity was plotted against the carbon number of each protein (Fig \ref{fig:Prod_POI}B).
The proteins with the highest productivity had the lowest carbon number whereas proteins with low productivity had higher carbon numbers.
This inverse qualitative trend was due to the fact that larger proteins require more amino acids and substrates to assemble them resulting in lower productivity.
\begin{figure}[t!]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/Prod_POI.pdf}
\caption{CFPS productivity performace for eight proteins for the control (black), AA uptake w/o synthesis (off white), and AA synthesis w/o uptake (grey). A. Productivity normalized to the specific glucose uptake rate (Error bars represent the 95\% CI of the ensemble). B. Mean productivity versus the carbon number for each corresponding protein.}
\label{fig:Prod_POI}
\end{figure}

\subsection{CFPS Carbon Yield}
Following the same outline as in examining the productivity, we calculated the carbon yield for each protein (Fig \ref{fig:Yield_POI}A).
The same trends followed, where the case without amino acid synthesis showed the highest carbon yield and the control case had comparable performance.
The third case (with no amino acid uptake) had the lowest yields; this is most likely because glucose is utilized to synthesize the necessary amino acids for each protein as well as power the system. 
To determine the carbon contribution from each substrate (glucose and amino acids) we examined the carbon flux going toward the production of deGFP for all three cases (Table \ref{tbl:yield_breakdown}).
For the first case, the system is relied on a mixture of glucose and some amino acids for the productin of deGFP with a carbon yield of 20.5\%.
Once amino acid synthesis is removed from the network (2nd case), each amino acid is utilized and the carbon yield increased to 22.2\%.
Only the necessary amount of amino acids were used for the production of deGFP, thus it may be hypothesized that all the glucose was used to power CFPS and does not contribute to the carbon yield.
If that is the case, the carbon yield without glucose contribution is 100\%, whereas for the first case its 92.5\%.
Finally for the third case (without amino acid supplemented), the carbon yield was reduced to 14.8\% and the system used about double the amount of glucose as the first two cases.
In this case glucose is being used to synthesize amino acids and provide the necessary energy requirements to power transcription and translation; this trend was seen across all proteins except for CAT.
Interestingly, CAT carbon yield showed similar performance for all three cases, however it still had a lower yield than the first two cases.
Each protein has different energy requirements for transcription and translation depending on its sequence.
Thus, for the case of CAT, energy requirements for transcription or translation were low enough that it's carbon yield was not hindered in the third case as significantly as for the other proteins.

\begin{table}[t!]
\centering
    \caption{Carbon contribution from glucose and each amino acid toward deGFP production for three cases: control, amino acid uptake without synthesis, and amino acid synthesis without uptake.}
    \renewcommand{\arraystretch}{0.9}
    \begin{tabular}{lccc} \toprule
        \parbox{3.2 cm}{Carbon \phantom{abcdefghi} Produced (mM)} & Control & \parbox{2.5 cm}{AA uptake \phantom{ai} w/o synthesis} & \parbox{2.5 cm}{AA synthesis w/o uptake} \\ \midrule
        deGFP & 9.8 & 9.6 & 9.9  \\ \toprule
        \parbox{3.2 cm}{Carbon \phantom{abcdefghi} Consumed (mM)} & \\ \midrule
        GLC & 37.3 & 33.7 & 66.9 \\
        ALA & 0.0 & 0.2 & -  \\
        ARG & 0.3 & 0.3 & - \\
        ASN & 0.5 & 0.4 & -  \\
        ASP & 0.4 & 0.6 & -  \\
        CYS & 0.2 & 0.1 & -  \\
        GLU & 2.1 & 0.6 & -  \\
        GLN & 0.4 & 0.3 & - \\
        GLY & 0.3 & 0.3 & -  \\
        HIS & 0.5 & 0.5 & - \\
        ILE & 0.0 & 0.6 & -  \\
        LEU & 1.0 & 1.0 & - \\
        LYS & 1.0 & 0.9 & - \\
		MET & 0.0 & 0.2 & -  \\
        PHE & 1.0 & 0.9 & - \\
        PRO & 0.4 & 0.4 & -  \\
        SER & 0.0 & 0.2 & - \\
        THR & 1.0 & 0.5 & -  \\
        TRP & 0.1 & 0.1 & -  \\
        TYR & 0.8 & 0.8 & - \\
        VAL & 0.6 & 0.7 & - \\ \hline
        Sum & 47.9 & 43.3 & 66.9  \\ \toprule
        Yield & 20.5\% & 22.2\% & 14.8\% \\
	Yield w/o GLC & 92.5\% & 100\% & - \\ \bottomrule
    \end{tabular}
\label{tbl:yield_breakdown}
\end{table}

We next investigated the effect of the carbon number of each protein to the carbon yield (Fig \ref{fig:Yield_POI}B).
The same inverse qualitative trend is observed as for productivity, however it was less significant.
The proteins with the lowest carbon number had the highest yield and the higher carbon number proteins had a lower carbon yield within each case. 
As the size of the protein to be expressed increases, the productivity and carbon yield decreases, thus proteins that are too large may not be feasible for cell-free production.
Thus, we wanted to examine the most sensitive parameters for cell-free productivity and carbon yield in order to optimize CFPS systems most effectively.
%Sequence specific flux balance analysis assumes a psuedo steady state, thus intermediate metabolites cannot be accumulated within the cell-free extract.
%In addition, ssFBA is solved by setting the production of the protein as the objective function.
%Therefore, carbon flux will travel throughout the network to optimize the flux through the protein synthesis reaction.
\begin{figure}[t!]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/Yield_POI.pdf}
\caption{CFPS carbon yield performace for eight proteins for the control (black), AA uptake w/o synthesis (off white), and AA synthesis w/o uptake (grey). A. Carbon yield (Error bars represent the 95\% CI of the ensemble). B. Mean carbon yield versus the carbon number for each corresponding protein.}
\label{fig:Yield_POI}
\end{figure}
